Galois Slicing as Differentiation

### 2025-07-11 main/language.tex
[ tex ] finish a sketch of the CBN translation

Let the signature
$\Sigma_{\mathsf{num}} = (\{\mathsf{num}\}, \{\mathrm{zero} : \tyUnit \to \mathsf{num}, \mathrm{add} :
\mathsf{num} \times \mathsf{num} \to \mathsf{num}\})$. This signature suffices to write the simple query
function in \exref{introduction-example}, where we interpret the $\mathrm{Label}$ type as the sum
$\tyUnit \tySum \tyUnit$ and the labels $\mathsf{a}$ and $\mathsf{b}$ as $\tmInl{\tmUnit}$ and
$\tmInr{\tmUnit}$. We consider several interpretations of $\Sigma_{\mathsf{num}}$ in $\Fam(\LatGal)$ and their
behaviour on the selection-and-sum query from \exref{introduction-example}. First, we note the type of the
reverse approximation map in this case. The type of the approximation maps depends on the input value. Our example input database was $\mathit{db} = [(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$, meaning that the type of the reverse approximation map for this database and any label $l$ is:
\begin{displaymath}
  \begin{array}{l}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \\
    \quad \partial\sem{\mathsf{num}}(\mathrm{query}\,(l,\mathit{db})) \multimap \mathbb{1} \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(0)) \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(1)) \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(1)) \times \mathbb{1}
  \end{array}
\end{displaymath}
in the category $\JoinSLat$, where $\partial\sem{\mathsf{num}}(x)$ is the lattice of approximations of the
number $x$ determined by our interpretation. In the codomain, the first four $\mathbb{1}$s correspond to the
positions of labels in the input, which we are not approximating, and the final $\mathbb{1}$ is the terminator
of the list. Note how, even if the $\partial\sem{\mathsf{num}}(x)$ does not depend on $x$ the type of the
output is still dependent on the shape of the input list: type dependency is used in a fundamental way in our
interpretation.

\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. We have
  $\partial\sem{\mathsf{num}}(x) = \mathbb{1}$ so the type of $(\partial \mathrm{query})_r$ is trivial.
\item We take $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. The
  type of the reverse map now becomes:
  \begin{displaymath}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \{\top,\bot\} \multimap \mathbb{1} \times (\mathbb{1} \times \{\top,\bot\}) \times (\mathbb{1} \times \{\top,\bot\}) \times (\mathbb{1} \times \{\top,\bot\}) \times \mathbb{1}
  \end{displaymath}
  where every position that corresponds to a number has been tagged with $\top$ for ``present'' and $\bot$ for
  ``not present''. This interpretation recovers the behaviour given in \exref{introduction-example}: running
  the reverse approximation map at the input ``$\mathsf{a}$'' at approximation $\top$ reveals that only the
  numbers in the rows tagged with ``$\mathsf{a}$'' in the input are used, and likewise for ``$\mathsf{b}$''.
\item Quantitative approximation information with non-trivial dependency can be obtained by using the interval
  approximation interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}. Recall that $\mathrm{query}\,(\mathsf{a},\mathit{db}) = 1$, so in the reverse
  direction we must choose an interval containing $1$ to discover the largest (i.e. least in the order)
  intervals that will give rise to this output as \emph{independent} changes to the input. For example, if we
  pick $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query})_r(\mathsf{a},\mathsf{db})([\frac{9}{10},\frac{11}{10}]) = \cdot, (\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}]), \cdot
  \end{displaymath}
  Thus, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$, either the first ``$\mathsf{a}$'' row
  could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in $[\frac{9}{10},\frac{11}{10}]$, and
  the number in the ``$\mathsf{b}$'' row is not relevant.
\end{enumerate}

@@

We illustrate the effect of the CBN translation on the query example from
\exref{introduction-example}. Applying $\cbn{-}$ to the type of the database

>>>>

Let the signature
$\Sigma_{\mathsf{num}} = (\{\mathsf{num}\}, \{\mathrm{zero} : \tyUnit \to \mathsf{num}, \mathrm{add} :
\mathsf{num} \times \mathsf{num} \to \mathsf{num}\})$. This signature suffices to write the simple query
function in \exref{introduction-example}, where we interpret the $\mathrm{Label}$ type as the sum
$\tyUnit \tySum \tyUnit$ and the labels $\mathsf{a}$ and $\mathsf{b}$ as $\tmInl{\tmUnit}$ and
$\tmInr{\tmUnit}$. We consider several interpretations of $\Sigma_{\mathsf{num}}$ in $\Fam(\LatGal)$ and their
behaviour on the selection-and-sum query from \exref{introduction-example}. First, we note the type of the
reverse approximation map in this case. The type of the approximation maps depends on the input value. Our example input database was $\mathit{db} = [(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$, meaning that the type of the reverse approximation map for this database and any label $l$ is:
\begin{displaymath}
  \begin{array}{l}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \\
    \quad \partial\sem{\mathsf{num}}(\mathrm{query}\,(l,\mathit{db})) \multimap \mathbb{1} \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(0)) \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(1)) \times (\mathbb{1} \times \partial\sem{\mathsf{num}}(1)) \times \mathbb{1}
  \end{array}
\end{displaymath}
in the category $\JoinSLat$, where $\partial\sem{\mathsf{num}}(x)$ is the lattice of approximations of the
number $x$ determined by our interpretation. In the codomain, the first four $\mathbb{1}$s correspond to the
positions of labels in the input, which we are not approximating, and the final $\mathbb{1}$ is the terminator
of the list. Note how, even if the $\partial\sem{\mathsf{num}}(x)$ does not depend on $x$ the type of the
output is still dependent on the shape of the input list: type dependency is used in a fundamental way in our
interpretation.
\newcommand{\TWO}{\mathbb{2}}

\begin{enumerate}
\item If we take $\sem{\mathsf{num}}_{\PrimTy} = \Disc(\mathbb{R})$, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ the embeddings of the usual zero and addition functions via $\Disc$, then the
  resulting interpretation contains no approximation information. We have
  $\partial\sem{\mathsf{num}}(x) = \mathbb{1}$ so the type of $(\partial \mathrm{query})_r$ is trivial.
\item We take $\sem{\mathsf{num}}_{\PrimTy} = \Lift(\Disc(\mathbb{R}))$, using the lifting monad from
  \secref{models-of-total-gps:lifting}, with $\sem{\mathrm{zero}}_{\PrimOp}$ and
  $\sem{\mathrm{add}}_{\PrimOp}$ defined from the unlifted interpretations above and the monad structure. The
  type of the reverse map now becomes, where $\TWO = \{\top.\bot\}$:
  \begin{equation}\label{eqn:reverse-query-approx-1}
    (\partial \mathrm{query})_r(l, \mathit{db}) : \TWO \multimap \mathbb{1} \times (\mathbb{1} \times \TWO) \times (\mathbb{1} \times \TWO) \times (\mathbb{1} \times \TWO) \times \mathbb{1}
  \end{equation}
  where every position that corresponds to a number has been tagged with $\top$ for ``present'' and $\bot$ for
  ``not present''. This interpretation recovers the behaviour given in \exref{introduction-example}: running
  the reverse approximation map at the input ``$\mathsf{a}$'' at approximation $\top$ reveals that only the
  numbers in the rows tagged with ``$\mathsf{a}$'' in the input are used, and likewise for ``$\mathsf{b}$''.
\item Quantitative approximation information with non-trivial dependency can be obtained by using the interval
  approximation interpretation from \exref{intervals-and-maxima-elements} and \secref{interval-approx}. We let
  $\sem{\mathsf{num}}_{\PrimTy} = \mathbb{R}_{\mathit{intv}}$ and interpret addition using the morphism given
  in \secref{interval-approx}. Recall that $\mathrm{query}\,(\mathsf{a},\mathit{db}) = 1$, so in the reverse
  direction we must choose an interval containing $1$ to discover the largest (i.e. least in the order)
  intervals that will give rise to this output as \emph{independent} changes to the input. For example, if we
  pick $[\frac{9}{10},\frac{11}{10}]$ as the interval, then:
  \begin{displaymath}
    (\partial\mathrm{query})_r(\mathsf{a},\mathsf{db})([\frac{9}{10},\frac{11}{10}]) = \cdot, (\cdot,[-\frac{1}{10},\frac{1}{10}]),(\cdot,\bot),(\cdot,[\frac{9}{10},\frac{11}{10}]), \cdot
  \end{displaymath}
  Thus, to achieve an output within $[\frac{9}{10},\frac{11}{10}]$, either the first ``$\mathsf{a}$'' row
  could be in $[-\frac{1}{10},\frac{1}{10}]$ or the second one could be in $[\frac{9}{10},\frac{11}{10}]$, and
  the number in the ``$\mathsf{b}$'' row is not relevant.
\end{enumerate}

@@

\newcommand{\Element}{(\TWO \times \TWO)}

We illustrate the effect of the CBN translation on the query example from
\exref{introduction-example}. Applying $\cbn{-}$ to the type of $\mathrm{query}$ yields:
\begin{displaymath}
  T(\mathrm{Label}) \times T(\tyList(T(T(\mathrm{Label}) \tyProd T(\mathsf{num})))) \to T(\mathsf{num})
\end{displaymath}
Thus, every substructure of the input and output has been annotated with usage information. Under
interpretation in $\Fam(\MeetSLat \times \JoinSLat^\op)$, the type of the reverse approximation map as a
morphism in $\JoinSLat$ at the input $\mathit{db}$ is now (suppressing some ``$\times \mathbb{1}$'' for
readability):
\begin{displaymath}
  (\partial \mathrm{query})_r(l, \mathit{db}) : \TWO \multimap \TWO \times (\TWO \times \Element \times \Element \times \Element)
\end{displaymath}
Comparing to the type in (\ref{eqn:reverse-query-approx-1}), we now gain much more fine-grained information on
which parts of the input are used. When $l = \mathsf{a}$, we have
$(\partial \mathrm{query})_r(\mathsf{a},\mathit{db})(\top) = (\top, (\top, (\top, \top), (\top, \bot), (\top,
\top)))$, indicating that the execution of this query had examined everything except the number in the second
entry of the database. With the previous interpretation, we did not have confirmation that the labels in each
row were actually required.
